---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
Read in the data:
```{r readcsv}
data <- read.csv("activity.csv")
summary(data)
```
Packages used for this assignment include 'dplyr','ggplot2',  'knitr' and 'kableExtra'.
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

## What is the mean total number of steps taken per day?
Exclude NA's and take sum of steps, grouped by date:
```{r histogram,message=FALSE}
data_2 <- data[complete.cases(data),]
data_tot <- data_2 %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps))
```

Histogram of total steps taken:
```{r}
hist(data_tot$tot_steps, main = "Total number of steps taken", xlab = "steps")
```
  
Mean number of steps:
```{r mean}
mean(data_tot$tot_steps)
```
  
Median number of steps:
```{r median}
median(data_tot$tot_steps)
```

## What is the average daily activity pattern?
Take average of steps per interval:
```{r timeseries, message=FALSE}
data_mean <- data_2 %>%
    group_by(interval) %>%
    summarise(steps = mean(steps))
```

Time series plot for average number of steps:
```{r}
with(data_mean, plot(interval, steps, type = "l"))
```
  
The maximum number of steps:
```{r}
max_step<- head(data_mean[with(data_mean, order(-steps)),], 3)
kable_classic(kable(max_step, format = "html"), full_width = F, position = "left", font_size = 18, html_font = "Arial")

```

Interval 835 has the maximum number of steps.

## Imputing missing values
Total number of rows with NA's:
```{r}
nrow(data[!complete.cases(data),])
```

For this assignment I decided to replace the NA's with the mean of the steps grouped by interval.

```{r, message= FALSE}
data3 <- data
data3 <- data3 %>%
    group_by(interval) %>%
    mutate(steps= ifelse(is.na(steps),
                               mean(steps, na.rm=TRUE), as.numeric(steps)))
```


Take sum of steps, grouped by date:
```{r, message=FALSE}
data_tot2 <- data3 %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps))
```

Histogram of total steps taken:
```{r}
hist(data_tot$tot_steps, main = "Total number of steps taken", xlab = "steps")
```
  
Mean number of steps:
```{r mean2}
mean(data_tot2$tot_steps)
```
  
Median number of steps:
```{r median2}
median(data_tot2$tot_steps)
```

Replacing the NA's with the mean of the number of steps per interval does not seem to change the mean, and the median is just slightly different.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day:
```{r, message=FALSE}
data3$date2 <- as.Date(data3$date)
data3 <- data3 %>%
    mutate(weekday = ifelse(weekdays(date2, abbreviate = T ) == "za" | 
                                weekdays(date2, abbreviate = T ) == "zo",
                            "weekend", "weekday"))
```

Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days:
```{r,message=FALSE}
data_mean2 <- data3 %>%
    group_by(interval, weekday) %>%
    summarise(mean_steps = mean(steps))
with(data_mean2, qplot(interval, mean_steps, facets = weekday~., geom = "line",
                       ylab = "Number of steps", xlab = "Interval"))
``` 
